---
title: "Census Data Quality for Policy Decisions"
subtitle: "Evaluating ACS Reliability for Algorithmic Decision-Making"
author: "Mohamad AlAbbas"
date: today
format:
  html:
    code-fold: true
    code-summary: "Show code"
    code-tools: true
    toc: true
    toc-depth: 3
    toc-location: left
    theme: cosmo
    embed-resources: true
execute:
  warning: false
  message: false
editor: visual
---

::: {.callout-note appearance="minimal"}
**Course:** MUSA 5080 — Geospatial Machine Learning for Urban Planning, University of Pennsylvania
**Portfolio:** [mohamadalabbas-phd.github.io](https://mohamadalabbas-phd.github.io)
**Summary:** Assesses the reliability of American Community Survey (ACS) estimates at county and tract level, identifying high-uncertainty areas where margin-of-error makes algorithmic policy decisions risky.
:::

# Setup

```{r setup}
# Load required packages (hint: you need tidycensus, tidyverse, and knitr)

library(tidycensus)
library(tidyverse)
library(knitr)
library(kableExtra)

# Set your Census API key

census_api_key("807ea1c0820a3e1e46dde3c53438622057fcc1ba")

# Choose your state for analysis - assign it to a variable called my_state

my_state <- "California"

# ── Visual Themes ─────────────────────────────────────────────────────────────
plotTheme <- theme(
  plot.title       = element_text(size = 14, face = "bold"),
  plot.subtitle    = element_text(size = 11, color = "grey40"),
  plot.caption     = element_text(size = 8,  color = "grey55", hjust = 1),
  axis.text        = element_text(size = 9),
  axis.title       = element_text(size = 10, face = "bold"),
  panel.grid.minor = element_blank(),
  panel.background = element_rect(fill = "#fafafa", color = NA),
  plot.background  = element_rect(fill = "white",   color = NA),
  legend.title     = element_text(size = 9,  face = "bold"),
  legend.text      = element_text(size = 8)
)

mapTheme <- theme_void() +
  theme(
    plot.title      = element_text(size = 14, face = "bold"),
    plot.subtitle   = element_text(size = 11, color = "grey40"),
    plot.caption    = element_text(size = 8,  color = "grey55", hjust = 1),
    legend.title    = element_text(size = 9,  face = "bold"),
    legend.text     = element_text(size = 8),
    plot.background = element_rect(fill = "white", color = NA)
  )

palette5 <- c("#f0f9e8", "#bae4bc", "#7bccc4", "#43a2ca", "#0868ac")
```

**State Selection:** I have chosen California for this analysis because: I am currently on working on the wildfires that had occurred there with a few partners from UC: San Diego, so I know a bit about the state and pop density + I wish to visit it during the winter break :)!

## Part 1: County-Level Resource Assessment

### 1.1 Data Retrieval

```{r county-data}
# Write your get_acs() code here
county_vars <- c( med_hh_income = "B19013_001", total_pop     = "B01003_001")

county_raw <- get_acs(geography = "county", state =  my_state, survey = "acs5", year = 2022, variables = county_vars, output = "wide")

# Clean the county names to remove state name and "County"
# Hint: use mutate() with str_remove()

county <- county_raw %>%
  mutate(
    county_name = str_remove(NAME, paste0(", ", my_state)),
    county_name = str_remove(county_name, " County$")
  ) %>%
  select(GEOID, county_name, med_hh_incomeE, med_hh_incomeM, total_popE, total_popM)

# Display the first few rows
head(county)
```

### 1.2 Data Quality Assessment

```{r income-reliability}
# Calculate MOE percentage and reliability categories using mutate()
county_reliability <- county %>%
  mutate(
    moe_percentage = round((med_hh_incomeM/med_hh_incomeE) * 100, 2),
    Reliability = case_when(
      moe_percentage < 5 ~ "High Confidence",
      moe_percentage >= 5 & moe_percentage <= 10 ~ "Moderate Confidence",
      moe_percentage > 10 ~ "Low Confidence"
    )
  )

# Create a summary showing count of counties in each reliability category
# Hint: use count() and mutate() to add percentages

reliability_summary <- county_reliability %>%
  count(Reliability, name = "Count") %>%
  mutate(Proportion = round(100 * Count / sum(Count), 1))

kable(reliability_summary, caption = "County Income Reliability Categories") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, position = "left")

```

### 1.3 High Uncertainty Counties

```{r high-uncertainty}
# Create table of top 5 counties by MOE percentage

top5 <- county_reliability %>%
  arrange(desc(moe_percentage)) %>%
  slice(1:5) %>%
  select(
    county_name,
    med_hh_incomeE,
    med_hh_incomeM,
    moe_percentage,
    Reliability
  )

# Format as table with kable() - include appropriate column names and caption

kable(
  top5,
  caption = "Top 5 Counties by Median Household Income MOE Percentage",
  col.names = c("County", "Median Income", "Margin of Error", "MOE %", "Reliability Category"),
  digits = 2,
  align = c("l", "c", "c", "c", "c"),
  format.args = list(big.mark = ",")
) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, position = "left")
```

**Data Quality Commentary:**

All five of these counties are among the lowest-density areas in California. Because their populations are so small, the ACS relies on limited samples to generate median income estimates, which introduces greater variability. This explains the large disparities and the relatively high margins of error (11–19%). As a result, algorithms that classify or rank counties using these figures could produce erroneous outcomes if they neglect the margins of error. For example, Alpine County appears to have a median income exceeding \$100,000, but its margin of error is more than $17,000 an uncertainty that is enormous relative to its ~1,000 residents. This is both a sampling size and representativeness issue, highlighting how misleading the raw point estimate can be without MOE context.

## Part 2: Neighborhood-Level Analysis

### 2.1 Focus Area Selection

```{r select-counties}
# Use filter() to select 2-3 counties from your county_reliability data
# Store the selected counties in a variable called selected_counties

selected_counties <- bind_rows(
  county_reliability %>%
    filter(Reliability == "High Confidence") %>%
    slice(1),
  county_reliability %>%
    filter(Reliability == "Moderate Confidence") %>%
    slice(1),
  county_reliability %>%
    filter(Reliability == "Low Confidence") %>%
    slice(1),
) %>%
  select(
    County = county_name,
    `Median Income` = med_hh_incomeE,
    `Margin of Error` = med_hh_incomeM,
    `MOE %` = moe_percentage,
    Reliability = Reliability,
    Population = total_popE
  )

# Display the selected counties with their key characteristics
# Show: county name, median income, MOE percentage, reliability category
kable(
  selected_counties,
  caption = "Largest County by Population in Each Reliability Category",
  align = c("l", "r", "r", "r", "l", "r"),
  format.args = list(big.mark = ",")
) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, position = "left")
```

**Comment on the output:** Because I specified no randomness in how the slice is sampling the data across the reliability categories, it quite literally picked the first match it had. Which is why all three are arranged alphabetically. On the positive side we still have atleast 1 sample from each category and Alpine is still with us :)! The lowest county by density in California.

### 2.2 Tract-Level Demographics

```{r tract-demographics}
# Define your race/ethnicity variables with descriptive names

race_vars <- c(
  total    = "B03002_001",
  white    = "B03002_003",
  black    = "B03002_004",
  hispanic = "B03002_012"
)


# Use get_acs() to retrieve tract-level data
# Hint: You may need to specify county codes in the county parameter

selected_geoid <- c("06001","06003","06005")  # Alameda, Alpine, Amador

county_codes <- stringr::str_sub(selected_geoid, 3, 5)  # -> "001","003","005"

tract_raw <- get_acs(
  geography = "tract",
  state     = my_state,
  county    = county_codes,
  survey    = "acs5",
  year      = 2022,
  variables = race_vars,
  output    = "wide"
)

# Calculate percentage of each group using mutate()
# Create percentages for white, Black, and Hispanic populations

tract_percent <- tract_raw %>%
  mutate(
    county_code  = substr(GEOID, 3, 5),
    pct_white    = 100 * (whiteE    / totalE),
    pct_black    = 100 * (blackE    / totalE),
    pct_hispanic = 100 * (hispanicE / totalE),
    # Make a readable tract label and remove state/country suffixes
    tract_label  = stringr::str_remove(NAME, paste0(", ", my_state)),
    tract_label  = stringr::str_remove(tract_label, ", United States$")
  ) %>%
  left_join(
    county %>%
      filter(GEOID %in% selected_geoid) %>%
      transmute(
        county_code = substr(GEOID, 3, 5),
        county_name
      ),
    by = "county_code"
  ) %>%
  select(
    GEOID, county_name, tract_label,
    totalE, whiteE, blackE, hispanicE,
    totalM, whiteM, blackM, hispanicM,
    pct_white, pct_black, pct_hispanic
  )

# Add readable tract and county name columns using str_extract() or similar

kable(
  tract_percent %>%
    select(
      County = county_name,
      Tract  = tract_label,
      `White (%)`    = pct_white,
      `Black (%)`    = pct_black,
      `Hispanic (%)` = pct_hispanic
    ) %>%
    arrange(County, Tract),
  caption = "Tract-Level Racial/Ethnic Composition in Selected Counties",
  col.names = c("County", "Tract", "White (%)", "Black (%)", "Hispanic (%)"),
  align = c("l","l","r","r","r"),
  format.args = list(big.mark = ",", digits = 1)
) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, position = "left")

```

### 2.3 Demographic Analysis

```{r demographic-analysis}
# Find the tract with the highest percentage of Hispanic/Latino residents
# Hint: use arrange() and slice() to get the top tract

top_hispanic_tract <- tract_percent %>%
  arrange(desc(pct_hispanic)) %>%
  slice(1) %>%
  select(GEOID, tract_label, county_name, pct_hispanic)

kable(top_hispanic_tract,
      caption = "Tract with Highest % Hispanic/Latino",
      col.names = c("GEOID", "Tract", "County", "Hispanic (%)")) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, position = "left")

# Calculate average demographics by county using group_by() and summarize()
# Show: number of tracts, average percentage for each racial/ethnic group

county_demo_avgs <- tract_percent %>%
  group_by(county_name) %>%
  summarise(
    "Number of Tracts" = n(),
    "Average White Percentage" = mean(pct_white, na.rm = TRUE),
    "Average Black Percentage" = mean(pct_black, na.rm = TRUE),
    "Average Hispanic Percentage"  = mean(pct_hispanic, na.rm = TRUE)
  )

# Create a nicely formatted table of your results using kable()

kable(
  county_demo_avgs,
  caption = "Average Tract Demographics by County",
  digits = 1,
  align = c("l","c","c","c","c"),
  col.names = c("County Names", "Number of Tracts", "Average White Percentage","Average Black Percentage","Average Hispanic Percentage")
) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, position = "left")

```

## Part 3: Comprehensive Data Quality Evaluation

### 3.1 MOE Analysis for Demographic Variables

```{r demographic-moe}
# Calculate MOE percentages for white, Black, and Hispanic variables
# Hint: use the same formula as before (margin/estimate * 100)

# Create a flag for tracts with high MOE on any demographic variable
# Use logical operators (| for OR) in an ifelse() statement

county_codes_state <- unique(stringr::str_sub(county$GEOID, 3, 5))

race_vars <- c(
  total    = "B03002_001",
  white    = "B03002_003",
  black    = "B03002_004",
  hispanic = "B03002_012"
)

tract_state_raw <- get_acs(
  geography = "tract",
  state     = my_state,
  county    = county_codes_state,  # all counties in the state
  survey    = "acs5",
  year      = 2022,
  variables = race_vars,
  output    = "wide"
)

tract_state_percent <- tract_state_raw %>%
  mutate(
    county_code  = substr(GEOID, 3, 5),
    pct_white    = 100 * (whiteE    / totalE),
    pct_black    = 100 * (blackE    / totalE),
    pct_hispanic = 100 * (hispanicE / totalE),
    tract_label  = stringr::str_remove(NAME, paste0(", ", my_state)),
    tract_label  = stringr::str_remove(tract_label, ", United States$")
  ) %>%
  left_join(
    county %>%
      transmute(
        county_code = substr(GEOID, 3, 5),
        county_name
      ),
    by = "county_code"
  )

tract_quality <- tract_state_percent %>%
  mutate(
    moe_total_pct    = 100 * (totalM    / totalE),
    moe_white_pct    = 100 * (whiteM    / whiteE),
    moe_black_pct    = 100 * (blackM    / blackE),
    moe_hispanic_pct = 100 * (hispanicM / hispanicE),
    high_moe_flag = (moe_white_pct > 15) | (moe_black_pct > 15) | (moe_hispanic_pct > 15)
  )

tract_quality_summary <- tract_quality %>%
  summarise(
    tracts_total     = n(),
    tracts_high_moe  = sum(high_moe_flag, na.rm = TRUE),
    percent_high_moe = round(100 * tracts_high_moe / tracts_total, 1)
  )

kable(
  tract_quality_summary,
  caption   = "Tract-Level High-MOE Summary (>15% on any demographic variable) — Statewide",
  col.names = c("Total Tracts", "High-MOE Tracts", "Percent High-MOE (%)"),
  align     = "c"
) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, position = "left")
```

### 3.2 Pattern Analysis

```{r pattern-analysis}
# Group tracts by whether they have high MOE issues
# Calculate average characteristics for each group:
# - population size, demographic percentages

pattern_table <- tract_quality %>%
  group_by(high_moe_flag, county_name) %>%
  summarise(
    tracts         = n(),
    avg_pop        = mean(totalE, na.rm = TRUE),
    avg_pct_white  = mean(pct_white, na.rm = TRUE),
    avg_pct_black  = mean(pct_black, na.rm = TRUE),
    avg_pct_hispanic = mean(pct_hispanic, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(high_moe_flag), county_name)

kable(
  pattern_table,
  caption = "Characteristics by High-MOE Status (Any Demographic Variable > 15% MOE) — Statewide",
  col.names = c(
    "High-MOE (Any >15%)",
    "County",
    "Tracts",
    "Avg Pop",
    "Avg White (%)",
    "Avg Black (%)",
    "Avg Hispanic (%)"
  ),
  align = c("c","l","r","r","r","r","r"),
  digits = c(NA, NA, 0, 0, 1, 1, 1),
  format.args = list(big.mark = ",")
) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, position = "left")

tract_flag_driver <- tract_quality %>%
  mutate(
    flag_white    = moe_white_pct    > 15,
    flag_black    = moe_black_pct    > 15,
    flag_hispanic = moe_hispanic_pct > 15,
    driver_groups = case_when(
      flag_white & !flag_black & !flag_hispanic ~ "White",
      !flag_white & flag_black & !flag_hispanic ~ "Black",
      !flag_white & !flag_black & flag_hispanic ~ "Hispanic",
      flag_white | flag_black | flag_hispanic   ~ "Multiple",
      TRUE                                       ~ "None"
    )
  )

driver_totals <- tract_flag_driver %>%
  filter(high_moe_flag) %>%
  summarise(
    White    = sum(flag_white,    na.rm = TRUE),
    Black    = sum(flag_black,    na.rm = TRUE),
    Hispanic = sum(flag_hispanic, na.rm = TRUE)
  ) %>%
  tidyr::pivot_longer(everything(),
                      names_to = "Group",
                      values_to = "Flagged Tracts")

kable(
  driver_totals %>% arrange(desc(`Flagged Tracts`)),
  caption   = "Which Groups Drove High-MOE Flags (MOE% > 15) — Statewide",
  col.names = c("Group", "Flagged Tracts"),
  align     = c("l","r"),
  format.args = list(big.mark = ",")
) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, position = "left")

```

**Pattern Analysis:** If we are to detect patterns using the instructions given, then 99.9% of all census tracts are within the MOE margins, meaning there is some type of data error present. That is a clear indication that the issue is effectively random at the tract level. However, collapsing the results by demographic group rather than by tract reveals that the burden of error is not evenly shared: it falls overwhelmingly on minority populations, with Black communities the most affected. This stems from well-known challenges of under-sampling in the ACS, which causes large variations within smaller racial categories. Tracts with high MOEs tend to have smaller populations overall, amplifying sampling error, and even within White-majority tracts the subgroup estimates for Black and Hispanic residents frequently exceed the 15% MOE threshold. In some cases, there are so few observations that subgroup estimates are unstable or missing altogether. Together, these dynamics show that while the pattern may look random at the tract level, the reliability problem is systematically tied to the representation of minority populations, raising clear concerns for algorithmic decision-making.



## Part 4: Policy Recommendations

### 4.1 Analysis Integration and Professional Summary

**Executive Summary:**

Across county- and tract-level analyses, two systematic patterns consistently appear. First, tracts and counties with smaller populations tend to have disproportionately high margins of error, making their estimates far less stable than those from larger areas. Second, the reliability of racial and ethnic subgroup estimates varies sharply: Black and Hispanic populations are much more likely to have margins of error above 15%, and in some cases, the ACS does not capture enough observations to produce valid estimates. Together, these patterns show that measurement error is pervasive but not random it reflects structural features of both tract size and demographic composition.

Communities facing the greatest risk of algorithmic bias are those that are either very small and rural or racially/ethnically diverse. Rural tracts, because of small sample sizes, may be flagged as unreliable and thus deprioritized in automated systems, despite having genuine needs. At the same time, urban minority communities, particularly those with large Hispanic or Black populations, often show the highest subgroup MOEs, meaning their conditions could be systematically misclassified or underestimated. In both cases, the communities already at risk of marginalization are the same ones where the data is least reliable.

The drivers of these problems are structural. In rural areas, small sample sizes inflate margins of error, while in diverse urban tracts, underrepresentation of minority subgroups disrupts the accuracy of need assessments. This underrepresentation is tied to long-standing stratification in data collection, where certain groups are less visible in surveys, and to socio-spatial self-selection, where minorities concentrate in particular neighborhoods that are often harder to measure with precision. These processes produce systematic biases: the very communities whose needs are greatest — low-income, minority, and geographically marginalized — are those most likely to be misrepresented in the data.

The Department should treat reliability as central to its algorithmic framework. Specifically, it should (a) adjust for MOE when prioritizing communities, so noisy estimates are not misclassified as real differences; (b) avoid strict cutoffs in low-confidence areas by using broader eligibility bands; (c) supplement ACS data with administrative or community-level sources in minority-dense neighborhoods where subgroup reliability is weakest; and (d) incorporate transparency and equity audits to ensure that stratification and data gaps do not reinforce existing inequalities. By embedding these safeguards, the Department can ensure its allocation strategies are both statistically sound and socially just.

### 4.2 Specific Recommendations

```{r recommendations-data}
# Create a summary table using your county reliability data
# Include: county name, median income, MOE percentage, reliability category

recommendations <- county_reliability %>%
  select(
    County = county_name,
    `Median Income` = med_hh_incomeE,
    `MOE %` = moe_percentage,
    `Reliability Category` = Reliability
  ) %>%
  mutate(
    Recommendation = case_when(
      `Reliability Category` == "High Confidence"     ~ "Safe for algorithmic decisions",
      `Reliability Category` == "Moderate Confidence" ~ "Use with caution – monitor outcomes",
      `Reliability Category` == "Low Confidence"      ~ "Requires manual review or additional data",
      TRUE                                            ~ NA_character_
    )
  )

# Add a new column with algorithm recommendations using case_when():
# - High Confidence: "Safe for algorithmic decisions"
# - Moderate Confidence: "Use with caution - monitor outcomes"
# - Low Confidence: "Requires manual review or additional data"

# Format as a professional table with kable()

kable(
  recommendations %>%
    arrange(Recommendation, County),
  caption = "Decision Framework for Algorithm Implementation (Arranged by Recommendation)",
  col.names = c("County", "Median Income", "MOE %", "Reliability Category", "Recommendation"),
  digits = 2,
  format.args = list(big.mark = ",")
) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, position = "left")


```

**Key Recommendations:**

1. **Counties suitable for immediate algorithmic implementation:** Alameda, Butte, Contra Costa, El Dorado, Fresno, Humboldt, Imperial, Kern, Kings, Lake, Los Angeles, Madera Marin, Mendocino, Merced, Monterey, Napa,  Nevada, Orange, Placer, Riverside, Sacramento, San Bernardino, San Diego, San Francisco, San Joaquin, San Luis Obispo, San Mateo, Santa Barbara, Santa Clara, Santa Cruz, Shasta,Siskiyou, Solano, Sonoma, Stainislaus, Sutter, Tulare, Ventura, Yolo, and Yuba

2. **Counties requiring additional oversight:** Amador, Calaveras, Colusa, Del Norte, Glenn, Inyo, Lassen, Mariposa, Modoc, San Benito, Tehama, and Tuolumne

3. **Counties needing alternative approaches:** Alpine, Mono, Plumas, Sierra, and Trinity

### 4.3 Questions for Further Investigation

1. Are high-MOE tracts clustered spatially (e.g., along rural–urban boundaries or in specific regions of the state), or do they appear evenly dispersed?

2. Do MOE patterns persist across ACS releases, or do they improve over time with larger samples? A time-series comparison could reveal whether underrepresentation of minority or rural communities is a persistent structural issue, similar to how you track flood or disaster impacts across years.

3. How do MOE patterns for racial and ethnic groups vary across states? Are high MOEs for Hispanic and Black populations a uniquely California phenomenon, or do they reflect a broader national issue embedded in ACS sampling design?
